---
title: "AGEC 936 - Module 4, Feb. 25"
author: "Micah Cameron-Harp<br>Kansas State University"
subtitle: "Difference-in-Differences (DID)<br>Two-way Fixed Effects (TWFE)"
output:
 slidy_presentation:
   font_adjustment: +1
   toc: false
   footer: "AGEC 936 - Micah Cameron-Harp"
bibliography: references.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(Statamarkdown)
```

# This week
* Difference-in-differences (DID) in potential outcome framework
* Two-way Fixed Effects (TWFE)
* Parallel trend assumption
* Staggered treatment timing: When can TWFE fail and some alternatives

# Potential outcome representation
* With a binary treatment, $d_i=\{0,\;1\}$, let us denote the outcome of interest as
$$
y_i(0)\;\text{if}\;d_i=0 \\
y_i(1)\;\text{if}\;d_i=1
$$
* With an *ideal* research design, the difference, 
$$
y_i(1)-y_i(0)
$$ 
can be interpreted as **causal effect** [@rubin1974estimating]. In practice, **do we observe both $y_i(0)$ and $y_i(1)$**? 
* The *ideal* average treatment (causal) effect is defined as
$$
E(y_i(1)-y_i(0))=E(y_{i}(1))-E(Y_{i}(0)).
$$
* Instead, we observe 
$$
y_i=y_i(0)+(y_i(1)-y_i(0))d_{i}.
$$
and can estimate
$$
E(y_{i}|d_i=1)-E(y_{i}|d_i=0).
$$
* What happens if $y_i(0)$, $y_i(1)$, and $d_i$ are not independent?
$$
E(y_{i}|d_i=1)-E(y_{i}|d_i=0) \\
=E(y_i(0)+(y_i(1)-y_i(0))|d_i=1)-E(y_i(0)|d_i=0) \\
=(E(y_i(1)|d_i=1)-E(y_i(0)|d_i=1))+(E(y_i(0)|d_i=1)-E(y_i(0)|d_i=0)).
$$
If the **Bias** term is non-zero, what should we do?

# Difference-in-differences (DID)
* **Set-up** 
Let's consider 2 x 2 case (two-period, $t={1,2}$, N-unit, $i={1,...,N}$, and two-group, $g={T,C}$ case). Potential outcomes are
$$
y_{it}(0)=\alpha_i+\gamma_t+\varepsilon_{it} \\
y_{it}(1)=\alpha_i+\gamma_t+\tau_i+\varepsilon_{it}
$$
For $i \in T$, $d_{i1}=0$ and $d_{i2}=1$ and for $i \in C$, $d_{i1}=0$ and $d_{i2}=0$.
* **Average Treatment Effect on Treated (ATT)**
$$
E(\tau_i|i \in T)
$$
This is normally what we want!
* **First differences**
$$
y_{i2}-y_{i1}=(\alpha_i+\gamma_2+\tau_i d_{i2}+\varepsilon_{i2}) - (\alpha_i+\gamma_1+\tau_i d_{i1}+\varepsilon_{i1}) \\
=(\gamma_2-\gamma_1)+\tau_i (d_{i2}-d_{i1})+(\varepsilon_{i2}-\varepsilon_{i1})
$$
Conditional expectations by group:
$$
E(y_{i2}-y_{i1}|i \in T)=(\gamma_2-\gamma_1)+E(\tau_i (d_{i2}-d_{i1})|i \in T)+E(\varepsilon_{i2}-\varepsilon_{i1}|i \in T) \\
=(\gamma_2-\gamma_1)+E(\tau_i |i \in T)
$$
$$
E(y_{i2}-y_{i1}|i \in C)=(\gamma_2-\gamma_1)+E(\tau_i (d_{i2}-d_{i1})|i \in C)+E(\varepsilon_{i2}-\varepsilon_{i1}|i \in C) \\
=(\gamma_2-\gamma_1)
$$
* **Difference-in-differences**
$$
E(y_{i2}-y_{i1}|i \in T)-E(y_{i2}-y_{i1}|i \in C)
=\{(\gamma_2-\gamma_1)+E(\tau_i |i \in T)\}-(\gamma_2-\gamma_1) \\
=E(\tau_i |i \in T)
$$

# Two-way Fixed Effects (TWFE)
* **Two-way Fixed Effects** We can also use the following regression equation:
$$
y_{it}=\alpha_i+\gamma_t+\tau_i d_{it}+\varepsilon_{it}
$$
With $T=2$, the OLS estimator for the equation above, denoted by $\hat{\tau}_{FE}$, is consistently estimating *ATT*. Need the following assumptions:
  + Strict exogeneity
  + Parallel trend 
* **"Parallel-trend" assumption**: Revisit the DinD expression,
$$
E(y_{i2}-y_{i1}|i \in T)-E(y_{i2}-y_{i1}|i \in C)
=\{(\gamma_2-\gamma_1)+E(\tau_i |i \in T)\}-(\gamma_2-\gamma_1) \\
=E(\tau_i |i \in T)
$$
This is valid if 
$$
\gamma_{it}=\gamma_{t}\;\text{for}\;\forall\;i.
$$

# A simulation: the equivalence between DID and TWFE
* **DGP**
$$
y_{it}=5+2(t-1)+4d_{it}+0.2x_{i}+\varepsilon_{it}\;\;\text{with}\;\;i=\{1,2,...,500\}\;\;\text{and}\;\;t=\{1,2\}
$$
$$
d_{i1}=0,\;\;
d_{i2}= \begin{cases}
1\;\;\text{if}\;\;i=\{1,...,250\} \\
0\;\;\text{otherwise}
\end{cases}
$$
$$
x_{i}=\begin{cases}
0.5+u_{i}\;\;\text{if}\;\;i=\{1,...,250\} \\
u_{i}\;\;\text{otherwise}
\end{cases}
\;\;\text{and}\;\;u_i \sim N(0,1)
$$

```{r simple, collectcode=TRUE}
library(lmtest)
library(sandwich)
# set seed
set.seed(11123)

# 500 units
N<-500
unit<-1:500

# Define treatment effect
tau<-4

# Treatment groups
d <- c(rep(0, N/2), rep(1, N/2))

# DGP
X <- d*0.5+rnorm(N)
# Pre
Y_1 <- 5+0.2*X+0.3*rnorm(N)
# Post
Y_2 <- 5+2+0.2*X+tau*d+0.3*rnorm(N)

# DinD
(mean(Y_2[d == 1]) - mean(Y_1[d == 1])) - (mean(Y_2[d == 0]) - mean(Y_1[d == 0]))
```

```{r simple2, collectcode=TRUE}
# Two-way Fixed Effects
# Data prep
dind_data <- data.frame("Y"=c(Y_1, Y_2),
                        "Treat"=c(rep(0,N), d),
                        "Unit"=c(unit, unit),
                        "Time"=c(rep(1,N), rep(2,N)))

# Two-way FE regression
TWFE <- lm(Y ~ Treat + as.factor(Time) + as.factor(Unit), data = dind_data)
Results<-coeftest(TWFE, vcov=vcovHC)
Results[2,]
```

**With $T=2$, DID and TWFE are numerically identical**

# Violation of parellel-trend assumption
* **DGP**
$$
y_{it}=5+2(t-1)x_{i}+4d_{it}+0.2x_{i}+\varepsilon_{it}\;\;\text{with}\;\;i=\{1,2,...,500\}\;\;\text{and}\;\;t=\{1,2\}
$$
$$
d_{i1}=0,\;\;
d_{i2}= \begin{cases}
1\;\;\text{if}\;\;i=\{1,...,250\} \\
0\;\;\text{otherwise}
\end{cases}
$$
$$
x_{i}=\begin{cases}
0.5+u_{i}\;\;\text{if}\;\;i=\{1,...,250\} \\
u_{i}\;\;\text{otherwise}
\end{cases}
\;\;\text{and}\;\;u_i \sim N(0,1)
$$

```{r simple_nonparellel}
library(lmtest)
library(sandwich)
# set seed
set.seed(11123)

# 500 units
N<-500
unit<-1:500

# Define treatment effect
tau<-4

# Treatment groups
d <- c(rep(0, N/2), rep(1, N/2))

# Adding the differential trend: 0.5*(t+1)*d
# DGP
X <- d*0.5+rnorm(N)
# Pre
Y_1 <- 5+0.2*X+0.3*rnorm(N)
# Post
Y_2 <- 5+2*X+0.2*X+tau*d+0.3*rnorm(N)

# DinD
(mean(Y_2[d == 1]) - mean(Y_1[d == 1])) - (mean(Y_2[d == 0]) - mean(Y_1[d == 0]))

# Two-way Fixed Effects
# Data prep
dind_data <- data.frame("Y"=c(Y_1, Y_2),
                        "Treat"=c(rep(0,N), d),
                        "Unit"=c(unit, unit),
                        "Time"=c(rep(1,N), rep(2,N)))

# Two-way FE regression
TWFE <- lm(Y ~ Treat + as.factor(Time) + as.factor(Unit), data = dind_data)
Results<-coeftest(TWFE, vcov=vcovHC)
Results[2,]
```

# What did we just estimate?
* Now, we no longer have
$$
\gamma_{it}=\gamma_{t}\;\text{for}\;\forall\;i.
$$
Thus,
$$
E(y_{i2}-y_{i1}|i \in T)-E(y_{i2}-y_{i1}|i \in C) \\
=E((\gamma_{i2}-\gamma_{i1})+\tau_i |i \in T)-E(\gamma_{i2}-\gamma_{i1}|i \in C) \\
=E(\tau_i|i \in T)+\underbrace{\{E(\gamma_{i2}-\gamma_{i1}|i \in T)-E(\gamma_{i2}-\gamma_{i1}|i \in C)\}}_{\text{Non-parallel trend bias}}.
$$
* In the previous example, we had
$$
\gamma_{it}=2(t-1)x_{i}
$$
which means
$$
\gamma_{i2}-\gamma_{i1}=2x_{i}.
$$
Therefore, the "non-parallel trend bias" term becomes,
$$
E(2x_{i}|i \in T)-E(2x_{i}|i \in C)=1
$$
since $E(x_{i}|i \in T)=0.5$ and $E(x_{i}|i \in C)=0$.

# What if we have $T>2$?
* **Single treatment timing**: We can test for parellel-trend/common-trend assumption if we have more than one "pre-treatment" periods. One way to do the pre-trend testing is to estimate
$$
y_{it}=\alpha_i+\gamma_t+\sum_t \delta_t \mathbb{1}_i (i \in T) + \varepsilon_{it}
$$
and "test" whether $\delta_t$ are jointly zeros for $t<t^*$ where $t^*$ is the treatment period. 
  + *Issues*: Can be under-powered and may lead to a biased estimate [@rothpre]
  + *Alternative approach*: "Honest" approach [@rambachan2019honest]
  + Functional forms? Theory?
* **Staggered treatment timing** and (dynamic) **heterogeneous treatment effects**: TWFE fails if the treatment is staggered AND the treatment effects are heterogeneous. Many recent papers discuss this issue. Two papers/approaches we will focus on
  + @de2020two
  + @callaway2020difference
